Relationships in the extremes and their influence on competition and coexistence

### 2022-03-02 paper/Paper.Rnw
adding to discussion and some edits

%Currently 150 of an allowed 150 words for Ecology Letters
\begin{abstract}
\noindent Species coexistence attracts wide 
interest in ecology. Modern coexistence
theory (MCT) identifies coexistence mechanisms,
one of which, storage effects, hinges on relationships between fluctuations
in environmental and competitive pressures. However, such relationships
are typically measured using covariance, which does not account for 
the possibility that environment and competition may be more related to each other
when they are strong than when weak, or vice versa. 
Recent work showed that such `asymmetric tail associations' (ATAs) are common
between ecological variables, and are important for 
extinction risk, ecosystem stability, and other phenomena. 
We extend the MCT, decomposing storage
effects to show the influence of ATAs. Analysis of a simple model
and an empirical example using diatoms
illustrate that ATA influences can be comparable in magnitude to other
mechanisms of coexistence, and that ATAs can make the 
difference between species coexistence and competitive exclusion.
ATA influences are an important new mechanism of coexistence.
\end{abstract}

@@

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it extinct \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``relative nonlinearity
in competition'' and ``storage effects'' \citep{Chesson_1994} which will be familiar to many readers. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2006, Adler_2010}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Descamps_2005, Caceres_1997, Narawani_2013}, and corals \citep{Alvarez-Noriega_2020}. 
Although MCT is mathematically complex to many ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

%Intro to ATAs
Every ecological discipline involves the study of relationships among biological and environmental variables,
but standard methods for examining the association between two variables provide only a small 
portion of the overall information content of the
relationship \citep{Ghosh_2020_AER,Ghosh_2020_Eco,Ghosh_2020_ecolevol,Ghosh_2021}. For instance, 
the three pairs of variables in Fig. \ref{fig:noise} all have the same
Pearson correlation, but nevertheless demonstrate very different types of association. Other metrics 
of association are also often used, in addition to correlations
(e.g., various forms of regression), but correlations are extremely common, and other
methods also reveal only a small portion of the information content of the relationship between variables \citep{Nelsen2006}.
The variables in Fig. \ref{fig:noise}a,c 
exhibit what we have previously referred to as ``asymmetric tail association'' [ATA; \cite{Ghosh_2020_AER}], i.e., 
they are more strongly related to each
other in the left (Fig. \ref{fig:noise}a, termed ``left-tail association'') or right (Fig. \ref{fig:noise}c,
termed ``right-tail association'') portions of the variables' distributions. Although ATAs 
and related ideas had previously been seldom 
considered in ecology (but see \citealp{deValpine_2014, Anderson_2019, Popovic_2019}), 
our recent work shows that ATAs are extremely common \citep{Ghosh_2020_AER}, and can influence 
many phenomena, including extinction risk \citep{Ghosh_2020_Eco}, Taylor's law \citep{Ghosh_2020_AER}, 
and stability of ecosystem functioning \citep{Ghosh_2020_ecolevol,Ghosh_2021}.
Although ATAs are now known to be important, the implications of ATAs for competition and species coexistence 
have, to our knowledge, not previously been considered; doing so is the overarching mission of this study.

%Intro to storage effects and why they might be influenced by ATAs
It is straightforward to see, intuitively, 
why ATAs may influence one mechanism of coexistence in particular: storage effects. We review, conceptually, 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and explain why ATAs may be relevant to them.
Storage effects allow for species to fluctuate over time while escaping extinction because storage effects provide a relative benefit to
whichever species is currently rare; this is the characteristic that makes storage effects a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effects can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance). 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. Third, a buffering mechanism must exist whereby species limit the influence of competition
in times when the environment is poor. Dormant stages such as seed banks are an
example buffering mechanism \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species is currently rare.
Importantly, storage effects can be quantified by measuring to what extent EC covariance contributes to differences between 
invader and resident growth rates \citep{Chesson_1994,Ellner_2016}.
Thus ATAs may, in principle, influence storage effects simply because storage effects hinge on EC covariation, and ATAs, when they 
occur between environment and competition variables, can be an aspect of EC covariation.

%***DAN: Go though and make sure all citation callouts list papers in date order, olderst to newest,
%except when multiple pubs by the same first author are listed they are all together.

@@

%Climate change may alter ATAs of coexistence-relevant ATAs
An additional motivation for quantifying the importance of ATAs for species coexistence is that there is reason to believe climate change may
alter ATAs of environmental variables which impact coexistence, as follows.
It is well known that climate change can alter the intensity, duration, extent and frequency of extreme environmental events, including
heat waves, floods, hurricanes and fires \citep{Meehl_2004, Jentsch_2007, Ummenhofer_2017, Lyon2019, Keelings2020}.  
If extreme events are also longer lasting, then the nature of the covariation between phenologically separated environmental signals 
may be modified by changes in extreme climatic events; in particular, ATAs between such variables may be altered. 
For instance, imagine a case in which early spring precipitation influences the growth of a plant species, A,
whereas late spring precipitation influences plant species B. If these species are competitors, then EC covariation
contributing to their coexistence, or failure to coexist, can boil down to covariation 
between early and late spring precipitation. Because longer lasting weather events are more likely to contribute
to both early and late spring rainfall variables, a tendency for spring rainfall events to become both more extreme and 
longer lasting increases the degree of right-tail association between early- and late-spring rainfall. 
Likewise, a tendency for spring droughts to become both more intense and longer lasting can increase
left-tail association between these variables. 
As we will show in this study, changes in ATAs can then
result in different competitive outcomes between species.
\cite{Chesson_2012} provides examples from an Australian grass-forb community and from an Arizona 
community of desert annuals that seem to have all the ingredients necessary to make effects of the
type we describe here a possibility.
To our knowledge, the potential importance of climatic changes in ATAs for species coexistence has 
not been studied, but the importance for coexistence of changes 
in means and variability of environmental variables has been studied in many systems 
[e.g., \cite{White_2001, Adler_2006, Jentsch_2007, Angert_2009, diPaola_2012, Parepa_2013, Descombes_2020, Usinowicz_2021}]. 
Changes in variability and in extreme events are distinct from the changes in ATAs we 
consider here, because ATAs pertain to \emph{relationships} between environmental and other
variables in the extremes, which are distinct from and can be altered independently
of changes in the extremes of the univariate marginal distributions themselves \citep{Nelsen2006,Ghosh_2020_AER}.

%make the above abstract ideas more concrete - 
%***DAN: This para commented out for length, added a sentence about the systems to the above paragraph to take the place of this text
%To make the above ideas more concrete, we briefly mention examples from prior work to 
%which the ideas may apply. 
%\cite{Chesson_2012} provides two examples of competitive plant communities in which different species
%are likely influenced differently by phenologically distict environmental variables: an Australian 
%grass-forb community, and an Arizona community of desert annuals. 
%The communities follow the criteria required for temporal storage effects to operate \citep{Chesson_2012}. In both 
%examples, distinct species have different periods of germination, competition is influenced by resource 
%consumption, and population growth is buffered by overlapping generations or via seed banks.
%Temporal differences between species in germination rates suggest the possibility that 
%population growth of distinct species may be driven by rainfall signals in distinct parts
%of the growing season. Thus all the ingredients seem to be present by which climatic changes may
%alter ATAs between environmental variables which are relevant for the coexistence
%of the example species, motivating our efforts, here, to perform a detailed study of how ATAs can influence competition.

%Review of lit on changes in means and variances of climate signals and associated impacts on coexistence  
%***DAN: this para commented out for length, added a couple of the sentences to an above para to substitute
%Although, to our knowledge, the potential importance of climatic changes in ATAs for species coexistence has 
%not been studied, the importance for coexistence of changes 
%in means and variability of environmental variables has been studied in many systems. We briefly 
%review some of that literature.
%\cite{diPaola_2012} concluded that coexistence of drought sensitive,
%flood resistant species with flood sensitive, drought resistant species in 
%Mediteranean coastal swamp forests of central Italy would be jeopardized
%by projected changes in mean rainfall. 
%\cite{Descombes_2020} found that if warming trends promote faster expansion toward 
%higher altitudes of herbivore populations than of plant populations, newly arrived
%herbivores at high altitudes can inflate plant species richness by reducing dominant plant species 
%in favor of subordinates. Contrastingly, \cite{Usinowicz_2021} concluded, also for
%alpine plants, that upward range expansion under warming trends would be good for the growth and survival of 
%all species they considered, but that changes in competition would more than offset these direct benefits. 
%Numerous studies take an ``event-focused'' 
%perspective \citep{Jentsch_2007}, examining how increased climatic variability, in the
%form of extreme weather events, can alter competitive interactions. For instance, 
%\cite{White_2001} showed that single extreme heat events can alter competition 
%between C3 and C4 plants. And \cite{Parepa_2013} showed that extreme variability in soil
%nutrient availability for plants, mediated by soil moisture content, can alter 
%the potential for invasive species to enter an ecosystem and establish, a process that involves
%the invader overcoming competition from resident species. 
%Using MCT, it has been demonstrated that increased variability can promote coexistence in plant 
%communities, via storage effects \citep{Adler_2006,  Angert_2009}.
%Changes in variability and in extreme events are distinct from the changes in ATAs we 
%consider here, because ATAs pertain to \emph{relationships} between environmental and other
%variables in the extremes, which are distinct from and can be altered independently
%of changes in the extremes of the univariate marginal distributions themselves \citep{Nelsen2006,Ghosh_2020_AER}.

@@

%Goal para
We previously demonstrated \citep{Ghosh_2020_AER} that ATAs are common in nature, and we argued above 
that it is plausible both that climate change may alter environmental ATAs, and that ATAs may influence species 
coexistence. Thus we seek to explore the consequences of ATAs for competition, coexistence, and the maintenance of 
biodiversity, by means of the following specific goals. 
G1) To augment MCT to show formally how ATAs play a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATAs contribute to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; 
spatial variation can be considered in future work.
G2) To explore, using the simple lottery model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other, previously known mechanisms of coexistence. This is to help 
illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
G3) To explore, using the lottery model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaningfully large.
G4) To provide an empirical example of species coexistence which highlights the role ATAs can play.
In the Discussion, we revisit the possibility that climate change will alter
ATAs of environmental variables and thereby influence competition and species coexistence.
%***DAN: Make sure we end up having such material in the Discussion
Overall, our study presents a new mechanism of species coexistence which may be influenced by climate change, 
and a means of understanding its theoretical and empirical importance.

@@

%***DAN: consider sending the lottery model components of this section to sup mat, for length reasons.

@@

For a general model, it is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environment-dependent factor $E_i(t)$ and as a decreasing function of a quantity representing
competitive pressure on species $i$, $C_i(t)$. We assume, for simplicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though this assumption can be circumvented, we make the assumption because it simplifies the
presentation of new ideas.

%***DAN: I think we should replace the terminology "invasion growth rate" and "IGR" by "growth when rare" and "GWR",
%throughout the paper. Has to be done in the main text and sup mat and the figures themselves (IGR is often labeled on 
%the figure panels themselves).

@@

\underline{summary - include 1-2 sentences about misnomer}
We produced and applied a mathematical extension to MCT that, for the first time, 
considers ATAs for the coexistence of species and maintenance of biodiversity. 
Implementing new methods of MCT \citep{Ellner_2019, Ellner_2016}, we provided a further tertiary 
decomposition to MCT and decompose the storage effect into two new terms, 
$\Delta_i^{[EC]}$, the contribution of ATAs, and $\Delta_i^{[E \nat C]}$, the contribution of correlation \emph{per se}. 
The results of our application to a simple lottery model and 
to a competitive diatom model show that ATA contributions to storage effects and to coexistence
are non-negligible and meaningfully contribute to the coexistence of species and thus the maintenance of biodiversity. 
Our results reinforce that common methods of correlation fail to reflect and include 
important information on the structure of the relationship between variables and amends
coexistence to the list of ecological theory and phenomena that ATAs have been found to 
be important for \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Though storage effects are typically described as the contribution
of covariance between drivers to coexistence, classically \emph{EC covariance},
\citep{Chesson_2012, Ellner_2016, Ellner_2019}, we suggest that storage effects, instead, 
be described as the contribution of the \emph{relationship} between the drivers.
Measures of covariance do not fully capture all aspects of associativity throughout the bivariate 
distribution of the variables; thus, it may be misleading to describe the contribution of the whole 
EC (or other drivers') relationship as the contribution of EC (or other drivers') covariance.
Accordingly, we have separated the mechanisms of storage effect 
into contributions of ATAs and contributions of correlation \emph{per se}.

@@

Our expansion allows for the parsing of storage effects into contribution from a pure symmetrical correlation and structures beyond symmetric correlation. Previous studies report storage effect as a contribution to coexistence of EC covariance; yet compute it as the contribution of the EC relationship (covariance and beyond; symmetric and asymmetric).

\underline{ATA specialization}
Not only is the contribution of ATA non-neglible but in some cases the presence or absence of ATAs 
can mediate the result of competition between species, i.e., coexistence or extirpation. 
For some parameters, the contribution of ATAs was the driving mechanism of 
coexistence – the IGR would be negative without it – deeming the scenario a case of 
ATA rescue. For other parameter values, ATA exclusion occurs where ATAs are the 
driving mechanism of the failure to coexist (exclusion) and the lack of ATAs would result in 
a positive IGR. Parameter
subspaces of ATA exclusion and ATA rescue suggest a new kind of specialization among species or \emph{ATA specialists} - species which only persist in habitat patches where ATAs contribute to its coexistence. 
%I think ata specialists are a thing - if their competitor is ubiquitous, they can only live in places on the landscape 
%where there are atas. Could have a discussion para on it, if it does not make the results. Or both. BTW, ATA exclusion has a 
%similar thing going on - a species excluded by ATAs (and competition) can only live on the portions of the landscape where
%there are no ATAs. Could also describe the temporal way of it - the ATA specialist can only survive in a era/epoch for which there
%are ATAs in its local, otherwise it'll have to go dormant or move elsewhere.

\underline{future work: study ATAs in environmental variables and how theyre changing - tie in with specialist}

Our study highlights the importance of ATAs for ecological phenomena 
and specifically for coexistence and the maintenance of biodiversity. 
The occurence of ATAs among environmental variables themselves, 
their contribution to coexistence, and their changes through space and time
can and should be explored further. 
The coupling of the magnitude of ATA contribution and environmental variability, 
as shown in our results and figures (Fig. \ref{fig:fig3} - \ref{fig:fig6}), 
suggests that ATAs and their effects
may become more important to consider for coexistence and possibly other 
ecological phenomena as environmental conditions evidently
become more variable.
As described in section \ref{section:introduction}, changes in the ATAs between environmental
variables may occur as the climate becomes more variable and extreme events become more 
expansive and extensive, causing relationships between separated events to become tighter at one end or another and thus asymmetric. Additionally, 
if ATA specialist exist, their persistance may
become at stake as the mechanism driving their coexistence changes with the climate.
Thus it is of interest to further study how tail associations of environmental variables are changing and how this might 
alter competition, coexistence, and the maintenance of biodiversity generally.
Our investigation brings to attention the consequences of such
possible \emph{relationships} in the extremes; however, 
simpler changes in extreme events themselves such as
intensity and frequency and their influence on coexistence also merit future work. 
Possible changes in tail association
patterns between environmental variables in nature can
influence the coexistence of species through the storage effect and not other mechanisms, but
extreme and changing climate is probable to affect other mechanims in other ways.
%In the Discussion, I intend to talk about how possible changes in tail association 
%patterns of env vars in nature may influence coexistence of species. It won't influence the 
%other mechanisms, though changes in the extreme values of environmental variables, known
%to be occurring, might. 
%Should also mention this second point in the Discussion as something that merits further research.
%though might be worth mentioning in the Discussion that events are getting more extreme. We have studied how *relationships*
%in the extremes influence coexistence, but probably worth studying (or cite papers if it's been done) how changes
%in the commonness and extremeness of extreme events themselves influence coexistence, and for that you'd have
%to also look at the other mechanisms. May still need a sup mat laying out all the math.

@@

\underline{conclude: how ur work answers/informs the broad question from the intro and lay out 2-3 key priorities for next research on this question}
We provided another showcase of the importance of ATAs in ecology, 
building on previous studies.
ATAs clearly have implications for coexistence and competition;
though more research is needed in exploring more complex models 
including spatiotemporal and multispecies models.
Hope to see this tertiary decomposition of MCT be applied to empirical research.
More investigations should be done to explore implications of ATAs 
to other ecological theories.

>>>>

%Currently 148 of an allowed 150 words for Ecology Letters
\begin{abstract}
\noindent Species coexistence attracts wide 
interest in ecology. Modern coexistence
theory (MCT) identifies mechanisms of coexistence,
one of which, storage effects, hinges on relationships between fluctuations
in environmental and competitive pressures. However, such relationships
are typically measured using covariance, which does not account for 
the possibility that environmental and competitive influences may be more related to each other
when they are strong than when weak, or vice versa. 
Recent work showed that such `asymmetric tail associations' (ATAs) are common
between ecological variables, and are important for 
extinction risk, ecosystem stability, and other phenomena. 
We extend the MCT to account for ATAs. Storage
effects are decomposed to show the influence of ATAs. Simulations of a simple model
and an empirical example using diatoms
illustrate that ATA influences can be comparable in magnitude to other
mechanisms of coexistence, and that ATAs can make the 
difference between species coexistence and competitive exclusion. 
\end{abstract}

@@

%Intro to competition, coexistence, MCT, etc.
Competition and the maintenance of biodiversity are research topics at the core of modern ecology 
\citep{MacArthur_1958, Hutchinson_1961, Schmida_1984}.
The competitive exclusion principle states 
that two species competing for the same limiting resource cannot coexist at constant densities - one will instead outcompete 
the other and drive it to extinction \citep{Hardin_1960}.
Modern coexistence theory [MCT; \cite{Chesson_1994,Chesson_2000,Barabas_2018}] is a 
framework for understanding how species nevertheless coexist, and biodiversity is maintained, in real ecosystems. 
Coexistence occurs through stabilizing factors and niche 
differentiation, in spite of the conclusions drawn for idealized scenarios by the competitive exclusion principle. 
MCT breaks down coexistence into the contributions of various mechanisms, with names such as ``relative nonlinearity
in competition'' and ``storage effects'' \citep{Chesson_1994} which will be familiar to many readers. MCT has been
applied to several systems, including annual \citep{Angert_2009, Hallett_2019} and perennial plants \citep{Adler_2010, Adler_2006}, 
tropical trees \citep{Usinowicz_2012}, plankton \citep{Narawani_2013, Descamps_2005, Caceres_1997}, and corals \citep{Alvarez-Noriega_2020}. 
Although MCT is mathematically complex to many ecologists, recent extensions make it more accessible \citep{Ellner_2016,Ellner_2019}.

%Intro to ATAs
Every ecological discipline involves the study of relationships among biological and environmental variables,
but standard methods for examining the association between two variables provide only a small 
portion of the overall information content of the
relationship \citep{Ghosh_2020_AER,Ghosh_2020_Eco,Ghosh_2020_ecolevol,Ghosh_2021}. For instance, 
the three pairs of variables in Fig. \ref{fig:noise} all have the same
Pearson correlation, but nevertheless demonstrate very different types of association. Other metrics 
of association are also often used, in addition to correlations
(e.g., various forms of regression), but correlations are extremely common, and other
methods also reveal only a small portion of the information content of the relationship between variables \citep{Nelsen2006}.
The variables in Fig. \ref{fig:noise}a,c 
exhibit what we have previously referred to as ``asymmetric tail association'' [ATA; \cite{Ghosh_2020_AER}], i.e., 
they are more strongly related to each
other in the left (Fig. \ref{fig:noise}a, termed ``left-tail association'') or right (Fig. \ref{fig:noise}c,
termed ``right-tail association'') portions of the variables' distributions. Although ATAs 
and related ideas had previously been seldom 
considered in ecology (but see \citealp{Anderson_2019, deValpine_2014, Popovic_2019}), 
our recent work has shown that, in fact, ATAs are extremely common in ecology \citep{Ghosh_2020_AER}, and can influence 
many ecological phenomena, including extinction risk \citep{Ghosh_2020_Eco}, Taylor's law \citep{Ghosh_2020_AER}, 
synchrony, and the temporal stability of ecosystem functioning \citep{Ghosh_2020_ecolevol,Ghosh_2021}.

%Intro to storage effects and why they might be influenced by ATAs
Although ATAs are now known to be common and important, the implications of ATAs for competition and species coexistence 
have, to our knowledge, not previously been considered; doing so is the overarching mission of this study. 
It is straightforward to see, intuitively, 
why ATAs may influence one mechanism of coexistence in particular: storage effects. We review, conceptually, 
what storage effects are \citep{Chesson_1994,Chesson_2012,Ellner_2016,Barabas_2018}, and we explain why ATAs may be relevant to them.
Storage effects allow for species to fluctuate over time while escaping extinction because storage effects provide a relative benefit to
whichever species is currently rare; this is the characteristic that makes storage effects a so-called \emph{stabilizing} mechanism 
of coexistence \citep{Chesson_2000}.
Storage effects can occur if three conditions are met \citep{Chesson_1994}. First, there must be positive 
covariation between the fluctuating quality of the environment (E)
for a species and the strength of competition (C) experienced by that species (called EC covariance). 
Second, growth responses to the environment must be species-specific, so that distinct species do not find the 
same environmental conditions to be optimal. Third, a buffering mechanism must exist whereby species limit the influence of competition
in times when the environment is poor. Dormant stages such as seed banks are an
example buffering mechanism \citep{Chesson_2012}. 
When present, these three factors combine so that EC covariance provides a relative benefit to whatever species are currently rare.
Importantly, storage effects can be quantified by measuring to what extent EC covariance contributes to differences between 
invader and resident growth rates \citep{Chesson_1994,Ellner_2016}.
Thus ATAs may, in principle, influence storage effects simply because storage effects hinge on EC covariation, and ATAs, when they 
occur between environment and competition variables, can be an aspect of EC covariation.

@@

%Climate change may alter ATAs of coexistence-relevant ATAs
An additional motivation for quantifying the importance of ATAs for species coexistence is that there is reason to believe climate change may
alter ATAs of environmental variables which impact coexistence, as follows.
It is well known that climate change can alter the intensity, duration, extent and frequency of extreme environmental events, including
heat waves, floods, hurricanes and fires \citep{Jentsch_2007, Meehl_2004, Ummenhofer_2017, Lyon2019, Keelings2020}.  
If extreme events are also longer lasting, then the nature of the covariation between phenologically separated environmental signals 
may be modified by changes in extreme climatic events; in particular, ATAs between such variables may be altered. 
For instance, imagine a case in which early spring precipitation influences the growth of a plant species, A,
whereas late spring precipitation influences plant species B. If these species are competitors, then EC covariation
contributing to their coexistence, or failure to coexist, can boil down to covariation 
between early and late spring precipitation. Because longer lasting weather events are more likely to contribute
to both early and late spring rainfall variables, a tendency for spring rainfall events to become both more extreme and 
longer lasting increases the degree of right-tail association between early- and late-spring rainfall. 
Likewise, a tendency for spring droughts to become both more intense and longer lasting can increase
left-tail association between these variables. As we will show in this study, changes in ATAs can then
result in different competitive outcomes between species. Although we focus here on temporal coexistence
mechanisms, ATAs between environmental variables measured in different places may also be altered 
by climate change, if more intensive climatic events are also more spatially extensive.

%make the above abstract ideas more concrete
To make the above ideas more concrete, we briefly mention examples from prior work to 
which the ideas may apply. 
\cite{Chesson_2012} provides two examples of competitive plant communities in which different species
are likely influenced differently by phenologically distict environmental variables: an Australian 
grass-forb community, and an Arizona community of desert annuals. 
The communities follow the criteria required for temporal storage effects to operate \citep{Chesson_2012}. In both 
examples, distinct species have different periods of germination, competition is influenced by resource 
consumption, and population growth is buffered by overlapping generations or via seed banks.
Temporal differences between species in germination rates suggest the possibility that 
population growth of distinct species may be driven by rainfall signals in distinct parts
of the growing season. Thus all the ingredients seem to be present by which climatic changes may
alter ATAs between environmental variables which are relevant for the coexistence
of the example species, motivating our efforts, here, to perform a detailed study of how ATAs can influence competition.

%Review of lit on changes in means and variances of climate signals and associated impacts on coexistence
Although, to our knowledge, the potential importance of climatic changes in ATAs for species coexistence has 
not been studied, the importance for coexistence of changes 
in means and variability of environmental variables has been studied in many systems. We briefly 
review some of that literature.
\cite{diPaola_2012} concluded that coexistence of drought sensitive,
flood resistant species with flood sensitive, drought resistant species in 
Mediteranean coastal swamp forests of central Italy would be jeopardized
by projected changes in mean rainfall. 
\cite{Descombes_2020} found that if warming trends promote faster expansion toward 
higher altitudes of herbivore populations than of plant populations, newly arrived
herbivores at high altitudes can inflate plant species richness by reducing dominant plant species 
in favor of subordinates. Contrastingly, \cite{Usinowicz_2021} concluded, also for
alpine plants, that upward range expansion under warming trends would be good for the growth and survival of 
all species they considered, but that changes in competition would more than offset these direct benefits. 
Numerous studies take an ``event-focused'' 
perspective \citep{Jentsch_2007}, examining how increased climatic variability, in the
form of extreme weather events, can alter competitive interactions. For instance, 
\cite{White_2001} showed that single extreme heat events can alter competition 
between C3 and C4 plants. And \cite{Parepa_2013} showed that extreme variability in soil
nutrient availability for plants, mediated by soil moisture content, can alter 
the potential for invasive species to enter an ecosystem and establish, a process that involves
the invader overcoming competition from resident species. 
Using MCT, it has been demonstrated that increased variability can promote coexistence in plant 
communities, via storage effects \citep{Adler_2006,  Angert_2009}.
Changes in variability and in extreme events are distinct from the changes in ATAs we 
consider here, because ATAs pertain to \emph{relationships} between environmental and other
variables in the extremes, which are distinct from and can be altered independently
of changes in the extremes of the univariate marginal distributions themselves \citep{Nelsen2006,Ghosh_2020_AER}.

@@

%Goal para
We previously demonstrated \citep{Ghosh_2020_AER} that ATAs are common in nature, and we argued above 
that it is plausible both that climate change may alter environmental ATAs, and that ATAs may influence species 
coexistence. Thus we seek to explore the consequences of ATAs for competition, coexistence, and the maintenance of 
biodiversity, by means of the following specific goals. 
G1) To augment MCT to show formally how ATA plays a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATA contributes to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; 
spatial variation can be considered in future work.
G2) To explore, using the simple lottery model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other, previously known mechanisms of coexistence. This is to help 
illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
G3) To explore, using the lottery model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaningfully large.
G4) To provide an empirical example of species coexistence which highlights the role ATAs can play.
In the Discussion, we revisit the possibility that climate change will alter
ATAs of environmental variables and thereby influence competition and species coexistence.
Overall, our study presents a new mechanism of species coexistence which may be influenced by climate change, 
and a means of understanding its theoretical importance and its importance for real ecosystems.

@@

@@

For a general model, it is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environment-dependent factor $E_i(t)$ and as a decreasing function of a quantity representing
competitive pressure on species $i$, $C_i(t)$. As for the lottery model example (Introduction), we here assume, for simplicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though this assumption can be circumvented, we make the assumption because it simplifies the
presentation of new ideas.

@@

We produced and applied a mathematical extension to MCT that, for the first time, 
considers ATAs for the coexistence of species and maintenance of biodiversity. 
Implementing new methods of MCT \citep{Ellner_2019, Ellner_2016}, we provided a further tertiary 
decomposition to MCT and decompose the storage effect into two new terms, 
$\Delta_i^{[EC]}$, the contribution of ATAs, and $\Delta_i^{[E \nat C]}$, the contribution of correlation \emph{per se}. 
The results of our application to a simple lottery model and 
to a competitive diatom model show that ATA contributions to storage effects and to coexistence
are non-negligible and meaningfully contribute to the coexistence of species and thus the maintenance of biodiversity. 
Our results reinforce that common methods of correlation fail to reflect and include 
important information on the structure of the relationship between variables and amends
coexistence to the list of ecological theory and phenomena that ATAs have been found to 
be important for \citep{Ghosh_2020_AER, Ghosh_2020_Eco, Ghosh_2020_ecolevol, Ghosh_2021}.  
Though storage effects are typically described as the contribution
of covariance between drivers to coexistence, classically \emph{EC covariance},
\citep{Chesson_2012, Ellner_2016, Ellner_2019}, we suggest that storage effects, instead, 
be described as the contribution of the \emph{relationship} between the drivers.
Measures of covariance do not fully capture all aspects of associativity throughout the bivariate 
distribution of the variables; thus, it may be misleading to describe the contribution of the whole 
EC (or other drivers') relationship as the contribution of EC (or other drivers') covariance.
Accordingly, we have separated the mechanisms of storage effect 
into contributions of ATAs and contributions of correlation \emph{per se}.

@@

%Our expansion allows for the parsing of storage effects into contribution from a pure symmetrical correlation and structures beyond symmetric correlation. Previous studies report storage effect as a contribution to coexistence of EC covariance; yet compute it as the contribution of the EC relationship (covariance and beyond; symmetric and asymmetric).

%\underline{ATA specialization}
Not only is the contribution of ATA non-neglible but in some cases the presence or absence of ATAs 
can mediate the result of competition between species, i.e., coexistence or extirpation. 
For some parameters, the contribution of ATAs was the driving mechanism of 
coexistence – the IGR would be negative without it – deeming the scenario a case of 
ATA rescue. For other parameter values, ATA exclusion occurs where ATAs are the 
driving mechanism of the failure to coexist (exclusion) and the lack of ATAs would result in 
a positive IGR. Parameter
subspaces of ATA exclusion and ATA rescue suggest a new kind of specialization among species or \emph{ATA specialists} - species which only persist in habitat patches where ATAs contribute to its coexistence. 
One can imagine a species whose coexistence with others is conditional on the asymmetric relationship between drivers of coexistence and whose populations may become excluded when ATAs erode. 
%I think ata specialists are a thing - if their competitor is ubiquitous, they can only live in places on the landscape 
%where there are atas. Could have a discussion para on it, if it does not make the results. Or both. BTW, ATA exclusion has a 
%similar thing going on - a species excluded by ATAs (and competition) can only live on the portions of the landscape where
%there are no ATAs. Could also describe the temporal way of it - the ATA specialist can only survive in a era/epoch for which there
%are ATAs in its local, otherwise it'll have to go dormant or move elsewhere.

%\underline{future work: study ATAs in environmental variables and how theyre changing - tie in with specialist}

Our study highlights the importance of ATAs for ecological phenomena 
and specifically for coexistence and the maintenance of biodiversity. 
The occurence of ATAs among environmental variables themselves, 
their contribution to coexistence, and their changes through space and time
can and should be explored further. 
The coupling of the magnitude of ATA contribution and environmental variability, 
as shown in our results and figures (Fig. \ref{fig:fig3} - \ref{fig:fig6}), 
suggests that ATAs and their effects
may become more important to consider for coexistence and possibly other 
ecological phenomena as environmental conditions evidently
become more variable.
As described in section \ref{section:introduction}, changes in the ATAs between environmental
variables may occur as the climate becomes more variable and extreme events become more 
expansive and extensive, causing relationships between separated events to become tighter at one end or another and thus asymmetric. Additionally, 
if ATA specialist exist, their persistance may
become at stake as the mechanism driving their coexistence changes with the climate.
Thus it is of interest to further study how tail associations of environmental variables are changing and how this might 
alter competition, coexistence, and the maintenance of biodiversity generally.
Our investigation brings to attention the consequences of such
possible \emph{relationships} in the extremes; however, 
simpler changes in [the marginal/ univariate distributions of] extreme events themselves such as
intensity and frequency and their influence on coexistence also merit future work. 
Possible changes in tail association
patterns between environmental variables in nature can
influence the coexistence of species through the storage effect and not other mechanisms, but
extreme and changing climate is probable to affect other mechanims in other ways.
%In the Discussion, I intend to talk about how possible changes in tail association 
%patterns of env vars in nature may influence coexistence of species. It won't influence the 
%other mechanisms, though changes in the extreme values of environmental variables, known
%to be occurring, might. 
%Should also mention this second point in the Discussion as something that merits further research.
%though might be worth mentioning in the Discussion that events are getting more extreme. We have studied how *relationships*
%in the extremes influence coexistence, but probably worth studying (or cite papers if it's been done) how changes
%in the commonness and extremeness of extreme events themselves influence coexistence, and for that you'd have
%to also look at the other mechanisms. May still need a sup mat laying out all the math.

@@

\underline{conclude **NOT NEEDED**: how ur work answers/informs the broad question from the intro and lay out 2-3 key priorities for next research on this question}
We provided another showcase of the importance of ATAs in ecology, 
building on previous studies.
ATAs clearly have implications for coexistence and competition;
though more research is needed in exploring more complex models 
including spatiotemporal and multispecies models.
Hope to see this tertiary decomposition of MCT be applied to empirical research.
More investigations should be done to explore implications of ATAs 
to other ecological theories.
